How to Avoid OOTA Without Really Trying

### 2024-09-03 paper/ootavacuous.tex
ootavacuous: Drop "rlx" subscripts from litmus tests
Woo-hoo!!!  They are are now all relaxed accesses, so no need to
differentiate any more.

Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

\begin{listing}[tbp]
% Constructed by hand.
{
\small
\begin{tabular}{r|l|l}
& \multicolumn{1}{c}{P0} & \multicolumn{1}{c}{P1} \\
\hline
1. &
\texttt{~~int r1 =\textsubscript{rlx} x} &
	\texttt{~~int r2 =\textsubscript{rlx} y} \\
2. &
\texttt{~~y =\textsubscript{rlx} r1} &
	\texttt{~~x =\textsubscript{rlx} r2} \\
\end{tabular}

@@

Listing~\ref{lst:Simple OOTA} shows a simple example where an OOTA
cycle might result in all of \co{x}, \co{y}, \co{r1}, and \co{r2}
having final values of 42, despite the fact that there is nothing in
the default-zero initial values or the executable code to support such
an outcome~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
The ``=\textsubscript{rlx}'' denotes a relaxed access, with variables
whose names are ``\co{r}'' followed by a number denoting registers,
and other names denoting shared variables.
First, P0's line~1 does a relaxed atomic load from \co{x} into \co{r1},
reading the value of P1's line-2 store rather than \co{x}'s initial
value of zero and somehow obtaining 42.
Second, P0's line~2 stores \co{r1}, and thus 42, to \co{y}.
Third, P1's line~1 loads 42 from \co{y} to \co{r2}.
Fourth and finally, P1's line~2 stores 42 to \co{x}, justifying the value
loaded by P0's line~1.

@@

\begin{listing}[tbp]
% Constructed by hand.
{
\small
\begin{tabular}{r|l|l}
& \multicolumn{1}{c}{P0} & \multicolumn{1}{c}{P1} \\
\hline
1. &
\texttt{~~int r1 =\textsubscript{rlx} x} &
	\texttt{~~int r2 =\textsubscript{rlx} y} \\
2. &
\texttt{~~y =\textsubscript{rlx} r1} &
	\texttt{~~x =\textsubscript{rlx} 42} \\
\end{tabular}

@@

This is a fine definition and is consistent with the words in the C++
standard, but it has a problem with intrathread rfi links
as exemplified by the following code: \\
% \begin{quote}
% Constructed by hand.
{
\small
\begin{tabular}{r|l}
% & \multicolumn{1}{c}{P0} \\
% \hline
1. &
\texttt{~~int r2 =\textsubscript{rlx} x} \\
2. &
\texttt{~~y =\textsubscript{rlx} r2} \\
3. &
\texttt{~~int r3 =\textsubscript{rlx} y} \\
4. &
\texttt{~~z =\textsubscript{rlx} r3} \\
\end{tabular}
} \\
% \end{quote}
This is an elaboration of P1 from
Listing~\ref{lst:Simple OOTA}
that adds \co{z} along with lines~3 and~4.
The problem is that a C++ implementation may
note that line~3 could well execute immediately after line~2, giving
other threads no chance to modify \co{y} in between.
Such an implementation might therefore behave as if line~3 had been
removed from the source code and line~4's \co{r3} had been replaced by
\co{r2}.
This means that line~4 stores the same value to \co{z} that was stored
to \co{y} by line~2.
And given that the load from \co{y} no longer exists, it cannot possibly
act as a temporal constraint.

@@

Some discussions of semantic dependencies assume that they
are strictly functions of the source code.
Although there are ways of making this work, many instances of
semantic dependency must be considered functions of particular executions.
Consider for example:
\begin{quote}
\begin{verbatim}
x = y * z;
\end{verbatim}
\end{quote}
(Here and below, we have written shared-variable accesses without
annotations, for brevity.  Please imagine they are all relaxed atomic.)

@@

Recall the Simple OOTA example in Listing~\ref{lst:Simple OOTA} on
page~\pageref{lst:Simple OOTA},
in which \co{thread1()} loads the value of \co{x} and stores it in
\co{y} while \co{thread2()} does the reverse.
A sufficiently perverse globally optimizing loose C++ compiler might
transform that program to the following before translating it into
machine code: \\
% Constructed by hand.
{
\small
\begin{tabular}{r|l|l}
% & \multicolumn{1}{c}{P0} & \multicolumn{1}{c}{P1} \\
% \hline
1. &
\texttt{~~int r1 = 42} &
	\texttt{~~int r2 = 42} \\
2. &
\texttt{~~y =\textsubscript{rlx} r1} &
	\texttt{~~z =\textsubscript{rlx} r2} \\
\end{tabular}
} \\
The loads previously on lines~6 and~12 have been replaced by constants.
Such a transformation complies with the loose C++ standard,
even though the resulting executable file would produce an unintuitive
OOTA outcome every time it runs!

>>>>

\begin{listing}[tbp]
% Constructed by hand.
{
\small
\begin{tabular}{r|l|l}
& \multicolumn{1}{c}{P0} & \multicolumn{1}{c}{P1} \\
\hline
1. &
\texttt{~~int r1 = x} &
	\texttt{~~int r2 = y} \\
2. &
\texttt{~~y = r1} &
	\texttt{~~x = r2} \\
\end{tabular}

@@

Listing~\ref{lst:Simple OOTA} shows a simple example where an OOTA
cycle might result in all of \co{x}, \co{y}, \co{r1}, and \co{r2}
having final values of 42, despite the fact that there is nothing in
the default-zero initial values or the executable code to support such
an outcome~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
Throughout this paper, variables whose names are ``\co{r}'' followed by
a number denote registers, and other names denote shared variables.
All accesses to shared variable are relaxed atomic accesses.
First, P0's line~1 does a relaxed atomic load from \co{x} into \co{r1},
reading the value of P1's line-2 store rather than \co{x}'s initial
value of zero and somehow obtaining 42.
Second, P0's line~2 stores \co{r1}, and thus 42, to \co{y}.
Third, P1's line~1 loads 42 from \co{y} to \co{r2}.
Fourth and finally, P1's line~2 stores 42 to \co{x}, justifying the value
loaded by P0's line~1.

@@

\begin{listing}[tbp]
% Constructed by hand.
{
\small
\begin{tabular}{r|l|l}
& \multicolumn{1}{c}{P0} & \multicolumn{1}{c}{P1} \\
\hline
1. &
\texttt{~~int r1 = x} &
	\texttt{~~int r2 = y} \\
2. &
\texttt{~~y = r1} &
	\texttt{~~x = 42} \\
\end{tabular}

@@

This is a fine definition and is consistent with the words in the C++
standard, but it has a problem with intrathread rfi links
as exemplified by the following code: \\
% \begin{quote}
% Constructed by hand.
{
\small
\begin{tabular}{r|l}
% & \multicolumn{1}{c}{P0} \\
% \hline
1. &
\texttt{~~int r2 = x} \\
2. &
\texttt{~~y = r2} \\
3. &
\texttt{~~int r3 = y} \\
4. &
\texttt{~~z = r3} \\
\end{tabular}
} \\
% \end{quote}
This is an elaboration of P1 from
Listing~\ref{lst:Simple OOTA}
that adds \co{z} along with lines~3 and~4.
The problem is that a C++ implementation may
note that line~3 could well execute immediately after line~2, giving
other threads no chance to modify \co{y} in between.
Such an implementation might therefore behave as if line~3 had been
removed from the source code and line~4's \co{r3} had been replaced by
\co{r2}.
This means that line~4 stores the same value to \co{z} that was stored
to \co{y} by line~2.
And given that the load from \co{y} no longer exists, it cannot possibly
act as a temporal constraint.

@@

Some discussions of semantic dependencies assume that they
are strictly functions of the source code.
Although there are ways of making this work, many instances of
semantic dependency must be considered functions of particular executions.
Consider for example:
\begin{quote}
\begin{verbatim}
x = y * z;
\end{verbatim}
\end{quote}

@@

Recall the Simple OOTA example in Listing~\ref{lst:Simple OOTA} on
page~\pageref{lst:Simple OOTA},
in which \co{thread1()} loads the value of \co{x} and stores it in
\co{y} while \co{thread2()} does the reverse.
A sufficiently perverse globally optimizing loose C++ compiler might
transform that program to the following before translating it into
machine code: \\
% Constructed by hand.
{
\small
\begin{tabular}{r|l|l}
% & \multicolumn{1}{c}{P0} & \multicolumn{1}{c}{P1} \\
% \hline
1. &
\texttt{~~int r1 = 42} &
	\texttt{~~int r2 = 42} \\
2. &
\texttt{~~y = r1} &
	\texttt{~~z = r2} \\
\end{tabular}
} \\
The loads previously on lines~6 and~12 have been replaced by constants.
Such a transformation complies with the loose C++ standard,
even though the resulting executable file would produce an unintuitive
OOTA outcome every time it runs!
